---
title: "OntologyMapper over extracted cMD values (R port of run_om.py)"
author: "MetaExtractor benchmark"
output:
  html_document:
    toc: true
    df_print: paged
params:
  # Directory containing this .rmd (the `harmonize/` dir). Mirrors run_om.py's HERE.
  harmonize_dir: "/Users/sehyunoh/OmicsMLRepo/MetaExtractor/benchmarks/cmd_sonnet5/harmonize"
  # Optional: python env that has `metaharmonizer` importable. Leave "" to use the
  # reticulate default (RETICULATE_PYTHON / active venv / conda).
  python: ""
  condaenv: ""
---

Build NCIt corpora and run `OntologyMapper` on the extracted disease / body_site /
treatment values from **both** model runs of the refined multi-seed pipeline
(`out_ms_refined/sonnet_r1/` and `out_ms_refined/haiku_r1/`; Section 3.3), writing
one crosswalk CSV per category.

Crosswalk columns: `query, ref_match, stage, match1` (top label), `match1_score`,
`match1_code` (obo_id joined from the built corpus).

Ontology roots (NCIt via EVSREST, keyless):

- `disease`   -> `NCIT:C2991`  (Disease or Disorder) — overrides the shipped
  oncology default `NCIT:C3262` Neoplasm, which does not cover cMD's
  general-disease vocabulary.
- `bodysite`  -> `NCIT:C32221` (Body Part) — registry default.
- `treatment` -> `NCIT:C1909`  (Pharmacologic Substance) — registry default.

> The staged mapping engine (`OntoMapEngine`) lives only in the Python
> `metaharmonizer` package, so it is called through **reticulate**. Everything
> else — JSON collection, dedup, corpus code-map join, CSV writing — is native R.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
library(reticulate)
library(jsonlite)
```

```{r paths}
HERE  <- normalizePath(params$harmonize_dir, mustWork = TRUE)
BENCH <- dirname(HERE)                                   # cmd_sonnet5/
MANI  <- jsonlite::fromJSON(file.path(BENCH, "manifest.json"),
                            simplifyVector = FALSE)

# extraction field -> OM corpus_category
CATS <- c(disease = "disease", body_site = "bodysite", treatment = "treatment")

# Extraction outputs to load: label -> dir under BENCH. Refined multi-seed run
# (Section 3.3 of the manuscript); swap the seed (r1/r2/r3) as needed. Each dir
# holds one <study>.json per manifest study, with per-sample enumeration.
RUN_DIRS <- c(sonnet = "out_ms_refined/sonnet_r1",
              haiku  = "out_ms_refined/haiku_r1")

# str -> atomic terms: normalise "/" to ";", split, trim, drop blanks
SPLIT <- function(s) {
  parts <- trimws(strsplit(gsub("/", ";", as.character(s)), ";", fixed = FALSE)[[1]])
  parts[nzchar(parts)]
}
```

```{r reticulate-env, eval=FALSE}
# Point reticulate at the interpreter that has `metaharmonizer`, if provided.
if (nzchar(params$condaenv)) use_condaenv(params$condaenv, required = TRUE)
if (nzchar(params$python))   use_python(params$python, required = TRUE)

# Override the shipped oncology disease root with general Disease-or-Disorder,
# mutating the live module dict (tuple key) before importing OntoMapEngine.
ome <- import("metaharmonizer.engine.ontology_mapping_engine", convert = FALSE)
reg <- py_get_attr(ome, "_CORPUS_REGISTRY")
py_set_item(reg, tuple("disease", "ncit"), "NCIT:C2991")

mh    <- import("metaharmonizer", convert = FALSE)
paths <- import("metaharmonizer._paths")
RETRIEVED_ONTOLOGIES_DIR <- py_str(paths$RETRIEVED_ONTOLOGIES_DIR)
```

```{r collect}
# ---- Full pre-harmonization snapshot --------------------------------------
# `MX` is the single object holding EVERY extracted cell across both model runs,
# before any MetaHarmonizer / OntologyMapper call. One row per atomic extracted
# value, keeping the {evidence, section, confidence} provenance that the query
# terms fed to OM ultimately discard. Columns:
#
#   source        "sonnet" | "haiku"    (which model run produced the cell)
#   study         manifest study id
#   level         "fields" (study-level) | "sample"
#   sample_index  1-based index into `samples` (NA for study-level fields)
#   field         extraction field name (disease, body_site, treatment, ...)
#   value         the extracted value ({value,...} cells unwrapped to .value)
#   evidence      evidence_quote from the cell, NA for bare-value cells
#   section       section the value was drawn from
#   confidence    model-reported confidence

# NULL-safe collapse of a (possibly list) scalar field to one string.
scal <- function(z) if (is.null(z)) NA_character_ else paste(unlist(z), collapse = "; ")

# One extraction cell -> rows(value, evidence, section, confidence). A
# {value, evidence_quote, section, confidence, ...} dict is unwrapped (keeping
# its metadata); a JSON array value yields one row per element.
cell_rows <- function(x) {
  ev <- se <- co <- NA_character_
  v  <- x
  if (is.list(x) && !is.null(names(x)) && "value" %in% names(x)) {
    ev <- scal(x$evidence_quote); se <- scal(x$section); co <- scal(x$confidence)
    v  <- x[["value"]]
  }
  xs   <- if (is.list(v)) v else list(v)
  rows <- lapply(xs, function(val) {
    if (is.null(val)) return(NULL)
    data.frame(value = as.character(val), evidence = ev, section = se,
               confidence = co, stringsAsFactors = FALSE)
  })
  rows <- rows[!vapply(rows, is.null, logical(1))]
  if (length(rows)) do.call(rbind, rows) else NULL
}

# Walk every study JSON in both output dirs into one long data.frame.
build_mx <- function() {
  rows <- list()
  for (label in names(RUN_DIRS)) {
    for (m in MANI) {
      p <- file.path(BENCH, RUN_DIRS[[label]], paste0(m$study, ".json"))
      if (!file.exists(p)) next
      d          <- jsonlite::fromJSON(p, simplifyVector = FALSE)
      fields     <- if (is.null(d$fields))  list() else d$fields
      samples    <- if (is.null(d$samples)) list() else d$samples
      containers <- c(list(fields), samples)
      levels     <- c("fields", rep("sample", length(samples)))
      idxs       <- c(NA_integer_, seq_along(samples))
      for (i in seq_along(containers)) {
        cont <- containers[[i]]
        if (!is.list(cont) || is.null(names(cont))) next
        for (field in names(cont)) {
          r <- cell_rows(cont[[field]])
          if (is.null(r)) next
          r$source       <- label
          r$study        <- m$study
          r$level        <- levels[i]
          r$sample_index <- idxs[i]
          r$field        <- field
          rows[[length(rows) + 1L]] <- r
        }
      }
    }
  }
  cols <- c("source", "study", "level", "sample_index", "field",
            "value", "evidence", "section", "confidence")
  if (!length(rows))
    return(setNames(data.frame(matrix(character(), 0, length(cols)),
                               stringsAsFactors = FALSE), cols))
  do.call(rbind, rows)[cols]
}

MX <- build_mx()
cat(sprintf("MX: %d extracted cells | %d studies | fields: %s\n",
            nrow(MX), length(unique(MX$study)),
            paste(sort(unique(MX$field)), collapse = ", ")))

# `collect()` now derives the atomic OM query terms for one field directly from
# MX, applying the same normalisation (SPLIT) and NA-token filtering as before,
# so MX is the single source of truth for what enters harmonization.
collect <- function(field) {
  v     <- MX$value[MX$field == field]
  bad   <- tolower(trimws(v)) %in% c("not_reported", "na", "nan", "")
  terms <- unlist(lapply(v[!bad], SPLIT), use.names = FALSE)
  sort(unique(terms))
}
```

```{r mx-tables}
# ---- MX -> per-study `samples x features` metadata tables ------------------
# Reshape the long MX snapshot into one wide table per study: rows = samples,
# columns = extraction fields, cells = extracted value. Study-level `fields`
# values (sample_index == NA) are broadcast across every sample row; a
# sample-level value for the same field takes precedence. Duplicate cells (a
# field that produced a JSON array) are collapsed with "; ". Studies with only
# study-level output become a single-row table.
#
# MetaExtractor-level cleaning applied here, before any harmonization:
#   * the extractor's `not_reported` sentinel (and blank cells) -> NA;
#   * columns that are NA for every sample are dropped (`sample_index` kept).
#
# Returns a named list keyed by study id; each element is a data.frame whose
# first column is `sample_index`, followed by one column per retained field.
mx_to_tables <- function(mx, source = "sonnet") {
  d <- mx[mx$source == source, , drop = FALSE]

  # Collapse duplicate cell values, treating `not_reported`/blank as missing:
  # drop those tokens first, and return NA if nothing meaningful remains.
  na_tokens <- c("not_reported", "na", "nan", "")
  join <- function(v) {
    v <- v[!is.na(v) & !(tolower(trimws(v)) %in% na_tokens)]
    if (!length(v)) NA_character_ else paste(unique(v), collapse = "; ")
  }
  studies <- unique(d$study)
  setNames(lapply(studies, function(st) {
    ds  <- d[d$study == st, , drop = FALSE]
    smp <- ds[ds$level == "sample", , drop = FALSE]     # per-sample cells
    fld <- ds[ds$level == "fields", , drop = FALSE]     # study-level constants
    features   <- unique(ds$field)
    sample_ids <- sort(unique(smp$sample_index))
    if (!length(sample_ids)) sample_ids <- NA_integer_  # study-level-only study

    # Lookups: field -> constant value; "index\rfield" -> sample value.
    const <- if (nrow(fld)) as.list(tapply(fld$value, fld$field, join)) else list()
    skey  <- if (nrow(smp))
      as.list(tapply(smp$value, paste(smp$sample_index, smp$field, sep = "\r"), join))
      else list()

    m <- matrix(NA_character_, length(sample_ids), length(features),
                dimnames = list(NULL, features))
    for (j in seq_along(features)) {
      f  <- features[j]
      cv <- const[[f]]
      for (i in seq_along(sample_ids)) {
        sv <- skey[[paste(sample_ids[i], f, sep = "\r")]]
        # sample value wins when present & non-NA, else fall back to the
        # broadcast study-level constant (also NA if it was not_reported).
        m[i, j] <- if (!is.null(sv) && !is.na(sv)) sv
                   else if (!is.null(cv) && !is.na(cv)) cv
                   else NA_character_
      }
    }
    out  <- data.frame(sample_index = sample_ids, m, check.names = FALSE,
                       stringsAsFactors = FALSE)
    # Drop all-NA columns (every value was not_reported/absent); keep the key.
    keep <- vapply(out, function(col) any(!is.na(col)), logical(1))
    keep[["sample_index"]] <- TRUE
    out[, keep, drop = FALSE]
  }), studies)
}

# One set of per-study tables per model run.
MX_TABLES       <- mx_to_tables(MX, source = "sonnet")  # primary run
MX_TABLES_HAIKU <- mx_to_tables(MX, source = "haiku")

for (nm in c("MX_TABLES", "MX_TABLES_HAIKU")) {
  tabs <- get(nm)
  cat(sprintf("%s: %d studies\n", nm, length(tabs)))
  for (st in names(tabs)) {
    tb <- tabs[[st]]
    cat(sprintf("  %-24s %3d samples x %d features\n", st, nrow(tb), ncol(tb) - 1L))
  }
}

# Optional: write one CSV per study per run alongside the crosswalks.
# runs <- list(sonnet = MX_TABLES, haiku = MX_TABLES_HAIKU)
# for (run in names(runs))
#   for (st in names(runs[[run]]))
#     write.csv(runs[[run]][[st]],
#               file.path(HERE, sprintf("mx_table_%s_%s.csv", run, st)),
#               row.names = FALSE)
```

```{r corpus-code-map}
# label(lower) -> obo_id, from the built corpus CSV.
corpus_code_map <- function(category) {
  csvp <- file.path(RETRIEVED_ONTOLOGIES_DIR, sprintf("ncit_%s_corpus.csv", category))
  df   <- read.csv(csvp, stringsAsFactors = FALSE, check.names = FALSE)
  cn   <- names(df)
  lab  <- if ("label"  %in% cn) "label"  else cn[grepl("label",        cn, ignore.case = TRUE)][1]
  oid  <- if ("obo_id" %in% cn) "obo_id" else cn[grepl("obo_id|code",  cn, ignore.case = TRUE)][1]
  setNames(as.character(df[[oid]]), tolower(trimws(as.character(df[[lab]]))))
}
```

```{r run, results="asis", eval=FALSE}
for (field in names(CATS)) {
  category <- CATS[[field]]
  terms <- collect(field)
  cat(sprintf("\n### %s (%s): %d unique atomic terms\n\n", field, category, length(terms)))
  cat("terms:", paste(terms, collapse = ", "), "\n\n")
  if (length(terms) == 0) next

  t0 <- Sys.time()
  eng <- mh$OntoMapEngine(
    corpus_category = category,
    query_ls        = as.list(terms),
    ontology_source = "ncit",
    s3_strategy     = NULL           # Stage 1 exact + Stage 2 embed + 2.5 synonym
  )
  res <- py_to_r(eng$run())          # pandas DataFrame -> data.frame
  cat(sprintf("OM done in %.0fs\n\n", as.numeric(difftime(Sys.time(), t0, units = "secs"))))

  codes <- corpus_code_map(category)
  res$match1_code <- vapply(res$match1, function(l) {
    v <- codes[tolower(trimws(as.character(l)))]
    if (length(v) == 0 || is.na(v)) "" else unname(v)
  }, character(1))

  cols <- intersect(c("query", "ref_match", "stage", "match1", "match1_score", "match1_code"),
                    names(res))
  out <- file.path(HERE, sprintf("crosswalk_%s.csv", field))
  write.csv(res[cols], out, row.names = FALSE)
  cat(sprintf("wrote %s\n\n", out))
}
```
